---
title: "Appendix - Statistics for Data Science Laboratory Project"
author: "Camilo Julian Berutti"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
editor: visual
engine: knitr
---

The databases analyzed in this project were downloaded with the GarminDB library (<https://github.com/tcgoetz/GarminDB>) in Python. These databases contain my Garmin Connect information about activities and other health-related metrics. It returns the following DBs:

-   fitbit.db

-   garmin.db

-   garmin_activities.db

-   garmin_monitoring.db

-   garmin_summary.db

-   summary.db

In this notebook I extract

```{r}
# Needed packages
#install.packages("DBI")
#install.packages("RSQLite")
library(DBI)
library(dplyr)
library(purrr)
# packages
library(lubridate)

```

```{r}
# Load environment variables
if (file.exists(".env")) {
  readRenviron(".env")
} else if (file.exists("../.env")) {
  readRenviron("../.env")
}

# conecta al archivo sqlite en tu .GarminDb
con_activities <- dbConnect(RSQLite::SQLite(), Sys.getenv("GARMIN_ACTIVITIES_DB"))

# ver qué tablas existen
#dbListTables(con_activities)
#dbGetQuery(con_activities, "SELECT * FROM activities LIMIT 5")

# traer datos a un data.frame
activities_summary <- dbReadTable(con_activities, "activities")
# eliminar columnas que son todo NA
activities_clean <- activities_summary %>% select(where(~ !all(is.na(.x))))
# Eliminar columnas que no me interesan
activities <- activities_clean %>% select(-"self_eval_effort", -"type")

dbDisconnect(con_activities) # Cierro DB
#head(activities)
```

```{r}
# conecta al archivo sqlite en tu .GarminDb
#dbDisconnect(con) # Cierro DB

con_garmin <- dbConnect(RSQLite::SQLite(), Sys.getenv("GARMIN_DB"))

# ver qué tablas existen
#dbListTables(con_garmin)
#dbGetQuery(con_garmin, "SELECT * FROM sleep WHERE day >= '2025-01-16' LIMIT 5")
```

Here I select the data I will analyze. And I export the useful dataframes as Rdata?.

Find sleep data in other DB (garmin.db, table sleep) , or in garmin_summary.db, table days_summary

maybe it is better if from daily_summary just bring the battery, and from the others, the sleep data. What about stress??

I will use garmin.db, garmin_activities.db. (Maybe garmin_monitoring.db)

### Cleaning garmin.db

```{r}
daily_summary_filtered <- dbGetQuery(con_garmin, "SELECT * FROM daily_summary WHERE day >= '2025-01-16'")

#print(daily_summary_filtered)
daily_summary <- daily_summary_filtered %>% select(where(~ !all(is.na(.x))))
daily_summary <- daily_summary %>% select(-"hydration_goal", -"floors_goal",-"hydration_intake",-"sweat_loss",-"calories_bmr",-"intensity_time_goal")
```

```{r}
sleep_events <- dbGetQuery(con_garmin, "SELECT * FROM sleep WHERE day >= '2025-01-16'")
#dbGetQuery(con_garmin, "SELECT * FROM stress LIMIT 5")
sleep_events<- sleep_events %>% select(where(~ !all(is.na(.x))))
```

```{r}
daily_summary <- daily_summary %>%
  mutate(
    day = as.Date(day),
    moderate_secs = as.numeric(as.difftime(moderate_activity_time, format="%H:%M:%OS", units="secs")),
    vigorous_secs = as.numeric(as.difftime(vigorous_activity_time, format="%H:%M:%OS", units="secs"))
  )# %>% select(-moderate_activity_time, -vigorous_activity_time)  # optional: drop original chr


sleep_events <- sleep_events %>%
  mutate(
    day   = as.Date(day),
    start = ymd_hms(start),
    end   = ymd_hms(end),
    total_sleep_secs = as.numeric(as.difftime(total_sleep, format="%H:%M:%OS", units="secs")),
    deep_sleep_secs  = as.numeric(as.difftime(deep_sleep,  format="%H:%M:%OS", units="secs")),
    light_sleep_secs = as.numeric(as.difftime(light_sleep, format="%H:%M:%OS", units="secs")),
    rem_sleep_secs   = as.numeric(as.difftime(rem_sleep,   format="%H:%M:%OS", units="secs")),
    awake_secs       = as.numeric(as.difftime(awake,       format="%H:%M:%OS", units="secs"))
  )

activities <- activities %>%
  mutate(
    start_time = lubridate::ymd_hms(start_time),
    stop_time  = lubridate::ymd_hms(stop_time),
    elapsed_secs = as.numeric(as.difftime(elapsed_time, format="%H:%M:%OS", units="secs")),
    moving_secs  = as.numeric(as.difftime(moving_time,  format="%H:%M:%OS", units="secs")),
    hrz_1_secs   = as.numeric(as.difftime(hrz_1_time,   format="%H:%M:%OS", units="secs")),
    hrz_2_secs   = as.numeric(as.difftime(hrz_2_time,   format="%H:%M:%OS", units="secs")),
    hrz_3_secs   = as.numeric(as.difftime(hrz_3_time,   format="%H:%M:%OS", units="secs")),
    hrz_4_secs   = as.numeric(as.difftime(hrz_4_time,   format="%H:%M:%OS", units="secs")),
    hrz_5_secs   = as.numeric(as.difftime(hrz_5_time,   format="%H:%M:%OS", units="secs"))
  )

activities$day <- as.Date(activities$start_time)
```

To characterize the sport contribution to the daily activities I calculate a new measure called "sport_exposure", which is calculated as: $$\text{sport_exposure} = \{0.6 * \text{avg_hr} + 0.4\text{max_hr}\} * \text{elapsed_secs}$$, the election of the weights and function form is completely arbitrary.

```{r}
# ---------------------------
# 1) Build the daily activity summary from raw activities, keeping a count of real activities.
#    We keep the raw sport_exposure as sport_exposure_raw (no imputation yet).
activity_load_raw <- activities %>%
  mutate(day = as.Date(day)) %>%
  mutate(
    combined_hr = 0.6 * as.numeric(avg_hr) + 0.4 * as.numeric(max_hr),
    sport_exposure_raw = combined_hr * as.numeric(elapsed_secs)
  ) %>%
  group_by(day) %>%
  summarise(
    sport_exposure_raw = sum(sport_exposure_raw, na.rm = TRUE),
    n_activities = n()                          # count of activity rows that day
  ) %>%
  ungroup()

# ensure sorted by date
activity_load_raw <- activity_load_raw %>% arrange(day)

# ---------------------------
# 2) Helper: imputation that searches ±10/20/30 days, applies penalties, fallback = 10% global mean
impute_local_mean <- function(current_day, data,
                              windows = c(10, 20, 30),
                              penalties = c(1.0, 0.5, 0.25),    # decreasing weight with distance
                              fallback_frac = 0.10) {            # 10% fallback of global mean
  # data is activity_load_raw
  for (i in seq_along(windows)) {
    days_window <- windows[i]
    penalty <- penalties[i]

    subset <- data %>%
      filter(!is.na(sport_exposure_raw),
             day >= current_day - days(days_window),
             day <= current_day + days(days_window))

    if (nrow(subset) > 0) {
      return(mean(subset$sport_exposure_raw, na.rm = TRUE) * penalty)
    }
  }
  # fallback if nothing found in any window
  global_mean <- mean(data$sport_exposure_raw, na.rm = TRUE)
  if (is.na(global_mean) || global_mean == 0) {
    return(0)      # extreme fallback if there's no data at all
  }
  return(fallback_frac * global_mean)
}

# ---------------------------
# 3) Join raw exposures into daily_summary (keeping n_activities and raw exposure)
daily_summary <- daily_summary %>%
  mutate(day = as.Date(day)) %>%
  left_join(
    activity_load_raw %>% select(day, sport_exposure_raw, n_activities),
    by = "day"
  )

# ---------------------------
# 4) Impute missing sport_exposure (use the local imputer). The imputed result we put in
#    a new column 'sport_exposure' (overwriting NA values).
daily_summary <- daily_summary %>%
  arrange(day) %>%
  mutate(
    # initial: if there's a real raw exposure (n_activities>0) use it, else NA (to impute)
    sport_exposure = ifelse(!is.na(sport_exposure_raw) & n_activities > 0,
                             sport_exposure_raw,
                             NA_real_)
  ) %>%
  mutate(
    sport_exposure = ifelse(
      is.na(sport_exposure),
      # map_dbl applies imputation for each day using activity_load_raw as reference
      map_dbl(day, ~ impute_local_mean(.x, activity_load_raw)),
      sport_exposure
    )
  )

# ---------------------------
# 5) Create an indicator whether the day had real activity (before imputation)
daily_summary <- daily_summary %>%
  mutate(had_real_activity = ifelse(!is.na(sport_exposure_raw) & n_activities > 0, TRUE, FALSE))

# ---------------------------
# 6) Define "very active" threshold based on the distribution of real activity days
#    We'll use the 75th percentile of sport_exposure_raw among days that had real activity.
very_active_threshold <- activity_load_raw %>%
  filter(!is.na(sport_exposure_raw) & sport_exposure_raw > 0) %>%
  pull(sport_exposure_raw) %>%
  quantile(probs = 0.75, na.rm = TRUE)

# If there were no real activity days, set a safe threshold = Inf so nothing becomes "very active"
if (length(very_active_threshold) == 0 || is.na(very_active_threshold)) {
  very_active_threshold <- Inf
}

# ---------------------------
# 7) Build the categorical variable activity_status according to your rules:
#    - "very active" if that day had real activity AND sport_exposure >= threshold
#    - "active"      if that day had real activity but below threshold
#    - "rest"        if that day had NO real activity but the day before was "very active"
#    - "inactive"    if that day and the day before had no exercise
#
# We compute a preliminary status for days with real activity, then use lag(prelim) to classify rest/inactive.
daily_summary <- daily_summary %>%
  arrange(day) %>%
  mutate(
    prelim_status = case_when(
      had_real_activity & sport_exposure >= very_active_threshold ~ "very active",
      had_real_activity & sport_exposure <  very_active_threshold ~ "active",
      TRUE ~ NA_character_
    )
  ) %>%
  # now finalize status using the previous day's prelim_status
  mutate(
    activity_status = case_when(
      !is.na(prelim_status) ~ prelim_status,
      is.na(prelim_status) & lag(prelim_status, default = NA) == "very active" ~ "rest",
      TRUE ~ "inactive"
    ),
    # make it an ordered factor if you want
    activity_status = factor(activity_status,
                             levels = c("very active", "active", "rest", "inactive"),
                             ordered = TRUE)
  ) %>%
  select(-prelim_status)   # drop helper

# ---------------------------
# 8) (Optional) Recompute normalized z-score if you want to include imputed values
daily_summary <- daily_summary %>%
  mutate(
    sport_exposure_z = ifelse(
      !is.na(sport_exposure),
      (sport_exposure - mean(sport_exposure, na.rm = TRUE)) / sd(sport_exposure, na.rm = TRUE),
      NA_real_
    )
  )

# done: daily_summary now contains:
#  - sport_exposure       (imputed where needed)

#  - had_real_activity    (TRUE/FALSE)
#  - activity_status      (very active | active | rest | inactive)
```

Remove undesired columns

```{r}
data <- full_join(daily_summary, sleep_events, by = "day")
data <- data %>%
  dplyr::select(-moderate_activity_time, 
         -vigorous_activity_time, 
         -total_sleep, 
         -deep_sleep, 
         -light_sleep, 
         -rem_sleep, 
         -awake,
         -sport_exposure_z,
         -sport_exposure_raw,
         -n_activities)
save(data,file = "Project_data.RData")
```

### Notes / rationale & tuning

-   I used the 75th percentile among *real activity days* as the threshold for **very active**. You can change `probs = 0.75` to another percentile (e.g. `0.80`) or use a fixed numeric threshold if you prefer.

-   The imputation windows are ±10/20/30 days with penalties `1`, `0.5`, `0.25` — i.e., close days contribute fully, more distant days are down-weighted. The global fallback is `10%` of the global mean if nothing is found.

-   `had_real_activity` is computed from the raw joined `sport_exposure_raw` and `n_activities` so we always know whether a day really had recorded activities before imputation.

-   `activity_status` uses the previous day's **prelim status** (i.e., whether the previous day actually had "very active") to determine whether a non-activity day is `rest` vs `inactive`, as you requested.

-   If you prefer “rest” to be triggered by the previous day being *either* "very active" OR "active", change the `lag(prelim_status) == "very active"` condition accordingly.

-   If you want, I can:

-   change the `very_active` threshold to a fixed numeric value (e.g., X seconds\*HR) instead of the 75th pctile, or

-   make the imputation use a smooth distance-weighted average (exponential/Gaussian kernel) instead of discrete windows and penalties — that gives a softer, continuous influence of nearby days.