---
title: "Statistics for Data Science - Laboratory Project"
author: "Camilo Julian Berutti"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
editor: visual
engine: knitr
---

```{r}
.libPaths("./library")
#install.packages("MVN")
#install.packages("lubridate")
#install.packages("tidyverse")
#install.packages("dplyr")
#install.packages("ggfortify")
#install.packages("corrplot")
#install.packages("psych")
#install.packages("ggplot2")
#install.packages("GGally"")
#install.packages("reshape2")
```

```{r}
#| message: false
#| warning: false
#| error: false
library(GGally)
library(ggplot2)
library(dplyr)
library(tidyr)
library(rlang)
library(readr)
library(MVN)
library(scales)
library(ggfortify)
library(ggrepel)
library(corrplot)
library(scatterplot3d)
library(caret)
library(cluster)
library(hopkins)
library("readr")
library(factoextra)
```

```{r}
load("Project_data.RData")
data
```

## About the Dataset

The dataset analyzed in this project originates from **Garmin Connect** exports and was processed with a custom Python extractor. The dataset consists in three CSV files combining information on **daily wellness measurements** (steps, heart rate, sleep, body battery, stress) and **physical activities** (running, swimming, etc.).

# Summary Analysis

```{r}
numeric_vars <- sapply(data, is.numeric)
numeric_col_names <- names(data)[numeric_vars]

results_list <- list()

normality_table <- data.frame(
  variable = character(),
  p_value = numeric(),
  normality = character(),
  stringsAsFactors = FALSE
)

for (col_name in numeric_col_names) {
  variable_data <- na.omit(data[[col_name]])

    cat("\n--- Plots for:", col_name, "---\n")
    op <- par(no.readonly = TRUE)
    par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

    hist(variable_data,
         main = paste("\nHistogram of", col_name),
         xlab = col_name)

    boxplot(variable_data,
            main = paste("\nBoxplot of", col_name),
            ylab = col_name)
    
    par(op)

    shapiro_test <- shapiro.test(variable_data)
    
    pval <- shapiro_test$p.value

    normality_result <- ifelse(pval < 0.05, "Non-normal", "Normal")
    
    cat("Shapiro-Wilk test for", col_name, "- p-value:",
        round(pval, 4), "\n")
    cat("Result:", normality_result, "\n")

    results_list[[col_name]] <- list(
      n = length(variable_data),
      mean = mean(variable_data),
      sd = sd(variable_data),
      p_value = pval,
      normality = normality_result
    )

    normality_table <- rbind(
      normality_table,
      data.frame(
        variable = col_name,
        p_value = pval,
        normality = normality_result,
        stringsAsFactors = FALSE
      )
    )
  }

  if (nrow(normality_table) > 0) {
    cat("\n\n================ NORMALITY SUMMARY TABLE ================\n")
    print(normality_table, row.names = FALSE)
    cat("==========================================================\n")
  }

  invisible(list(
    summary_table = normality_table,
    variable_summaries = results_list
  ))
```

### Curiosity bonus - period segmentation

```{r}
normality_segmented = read.csv("normality_results.csv") #created in Berutti_Project.Rmd
```

## Correlation

As the data is not normal, we can´t use Pearson method, we use Spearman :

```{r}
data_clean <- data

exclude_cols <- c("day", "activity_status", "had_real_activity", "start", "end")

numeric_for_corr <- data_clean %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select(-any_of(exclude_cols))

corr_matrix <- cor(numeric_for_corr, use = "pairwise.complete.obs")

corrplot(corr_matrix, cor_method = "spearman", method = "color", tl.cex = 0.6, order = "hclust") 
```

# Principal Component Analysis

We have a lot of variables, therefore it could be useful to analyze if we can find a new reference system (via a linear combination of the original variables) that allow us to explain the variance of the data.

```{r}
data_clean_complete <- na.omit(data_clean) 
original_rows <- nrow(data_clean)
cleaned_rows <- nrow(data_clean_complete)
cat("Removed", original_rows - cleaned_rows, "rows containing NA values.\n")
col_remov_number = c(1,17,18,19,20)
set.seed(123)


#data_clean_complete <- data_clean IT affect the results

train_index <- createDataPartition(data_clean_complete$had_real_activity, p = 0.7, list = FALSE)
train_data <- data_clean_complete[train_index, ]
test_data  <- data_clean_complete[-train_index, ]

train_scaled <- scale(train_data[,-col_remov_number])
test_scaled  <- scale(test_data[,-col_remov_number], center = attr(train_scaled, "scaled:center"),
scale = attr(train_scaled, "scaled:scale"))

pca <- prcomp(train_scaled, center = T, scale. = T)


```

We have a $\boldsymbol{X} = (X_1, X_2, \dots, X_p)^\top$: $p$-dimensional random vector (variable train_scaled) with mean vector $\boldsymbol{\mu}$ and positive definite covariance matrix $\boldsymbol{\Sigma}$. Elements are usually correlated.

By using PCA we replace $\boldsymbol{X}$ by a vector $\boldsymbol{Y} = (Y_1, Y_2, \dots, Y_p)^\top$ called **principal component (PC) vector of** $\boldsymbol{X}$ such that the components $Y_j$ are uncorrelated linear combinations of $X_1, X_2, \dots, X_p$.

```{r}
summary(pca)$importance[, 1:6]
```

```{r}
pca$rotation[,1:3]
```

```{r}
loadings_matrix <- pca$rotation[, 1:3]
threshold <- 0.3
loadings_relevant <- apply(loadings_matrix, 2, function(x) {
  x[abs(x) < threshold] <- NA
  return(x)
})

loadings_relevant
```

We can see that the components can describe

**PC1: Recovery vs. Stress:**

-   **High score:** Represents a day with good sleep quality.

-   **High negative score:** Represents a day with **high stress**.

**PC2: High intensity and caloric burn:**

-   **High negative score:** Since the loadings are negative, a *low* (more negative) score on PC2 actually means **high activity**. It represents a day with a hard workout (vigorous activity), high heart rate, and massive calorie burn.

-   **Low negative score:** Would represent a sedentary day with low caloric expenditure.

**PC3: Low vs. high intensity exercise:**

This component differentiate between general daily movement (walking) and specific sport exposure.

-   **High negative score:** Represents a day defined by **volume and walking**. You hit a high step count and recharged your battery (perhaps a low-stress, active recovery walking day).

-   **High positive score:** Represents a day defined by **specific sports** or moderate activity that might not be step-heavy (like cycling, swimming, or weight training) captured by `sport_exposure`.

```{r}
p_biplot <- autoplot(pca,loadings = TRUE,loadings.colour = "red",loadings.label = TRUE,loadings.label.repel = TRUE,loadings.label.size = 3,loadings.label.color = "red",alpha = 0.4,size = 0.8)
plot(p_biplot)
```

```{r}
library(plotly)

scale_factor <- 5 

scores_data <- as.data.frame(pca$x[, 1:3])
scores_data$Observation <- 1:nrow(scores_data)

loadings_data <- as.data.frame(pca$rotation[, 1:3])
loadings_data$Variable <- rownames(loadings_data)

loadings_scaled <- loadings_data
loadings_scaled[, 1:3] <- loadings_data[, 1:3] * scale_factor


p <- plot_ly()

p <- add_trace(p, 
               data = scores_data, 
               x = ~PC1, y = ~PC2, z = ~PC3,
               type = "scatter3d", 
               mode = "markers",
               marker = list(size = 3, opacity = 0.5, color = 'blue'),
               text = ~paste("Obs:", Observation),
               name = "Observations",
               inherit = FALSE) 


p <- add_text(p, 
              data = loadings_scaled,
              x = ~PC1, y = ~PC2, z = ~PC3,
              text = ~Variable,
              textfont = list(color = 'red', size = 10),
              name = "Variables",
              inherit = FALSE) 

for (i in 1:nrow(loadings_scaled)) {
  p <- add_trace(p, 
                 x = c(0, loadings_scaled$PC1[i]), 
                 y = c(0, loadings_scaled$PC2[i]), 
                 z = c(0, loadings_scaled$PC3[i]), 
                 type = "scatter3d",
                 mode = "lines",
                 line = list(color = 'red', width = 3),
                 showlegend = FALSE,
                 hoverinfo = "none",
                 inherit = FALSE) 
}

p <- p %>% layout(
  title = "3D PCA Biplot",
  scene = list(
    xaxis = list(title = "PC1"),
    yaxis = list(title = "PC2"),
    zaxis = list(title = "PC3")
  )
)

p



```

# Logistic Regression over PCA data

We want to see if it is possible to classify between days with/without sport activities. We have the categorical variable had_real_activity, that has TRUE or FALSE as possible values.

First I will spoil the use of the LR model, to find the optimal number of PC to consider, but I will show it in more detail later.

```{r}
train_pca_all <- as.data.frame(pca$x)
test_pca_all  <- as.data.frame(predict(pca, newdata = test_scaled))

acc_results <- data.frame(Components = integer(), Accuracy = numeric())

max_components_to_consider <- 9

for (k in 1:max_components_to_consider) {
  
  train_pca <- train_pca_all[, 1:k, drop = FALSE]
  test_pca  <- test_pca_all[, 1:k, drop = FALSE]

  train_pca$had_real_activity <- train_data$had_real_activity
  test_pca$had_real_activity  <- test_data$had_real_activity

  model <- glm(had_real_activity ~ ., data = train_pca, family = binomial)

  test_pca$pred_prob <- predict(model, newdata = test_pca, type = "response")
  test_pca$pred_class <- test_pca$pred_prob > 0.5

  # You no longer need 'na.rm = TRUE' because there are no NAs
  acc <- mean(test_pca$pred_class == test_pca$had_real_activity) 
  acc_results <- rbind(acc_results, data.frame(Components = k, Accuracy = acc))
}

```

```{r}
ggplot(acc_results, aes(x = Components, y = Accuracy)) +
  geom_point(color = "blue", size = 2) +
  geom_line(color = "blue", alpha = 0.6) +
  labs(
    title = "Logistic regression accuracy\nas a function of #PCA components",
    x = "Number of principal components",
    y = "Accuracy (test set)"
  ) +
  theme_minimal(base_size = 14)
```

Logistic Regression:

```{r}
k= 3 # Number of components to keep

train_pca <- train_pca_all[, 1:k, drop = FALSE] # THIS WILL BE MY DATA NOW
test_pca  <- test_pca_all[, 1:k, drop = FALSE]

train_pca$had_real_activity <- train_data$had_real_activity
test_pca$had_real_activity  <- test_data$had_real_activity

data_log_reg <- glm(had_real_activity ~ ., data = train_pca, family = binomial)
```

```{r}
summary(data_log_reg) # PC1 recovery,PC2  high intensity,PC3  low intensity
```

We check if the dataset is balanced:

```{r}
table(train_pca$had_real_activity)
```

We see that the amount of values is similar for both classes. To know which is the reference class :

```{r}
contrasts(train_pca$had_real_activity)
```

Therefore FALSE is the reference class. (Non-active day)

```{r}
#train_pca
train_pca$pred_prob <- predict(data_log_reg, newdata = train_pca, type = "response")
train_pca$pred_class <- ifelse(train_pca$pred_prob > 0.5, "TRUE", "FALSE")

conf_train <- table(Predicted = train_pca$pred_class, Actual = train_pca$had_real_activity)

conf_train
acc_train <- sum(diag(conf_train)) / sum(conf_train)
cat("Training accuracy:", round(acc_train, 3), "\n\n")
```

Now we check the accuracy in the test set:

```{r}
test_pca$pred_prob <- predict(data_log_reg, newdata = test_pca, type = "response")
test_pca$pred_class <- ifelse(test_pca$pred_prob > 0.5, "TRUE", "FALSE")
conf_test <- table(Predicted = test_pca$pred_class, Actual = test_pca$had_real_activity)
conf_test
acc_test <- sum(diag(conf_test)) / sum(conf_test)
cat("\nTest accuracy:", round(acc_test, 3), "\n")
```

For some mysterious reason the accuracy in the test set is greater than in the training. (with this dataset size misidentifying the class of one sample corresponds to ¬1.4% therefore it might be only by chance)

------------------------------------------------------------------------

# Exploratory Factor Analysis

We have a p-dimensional vector $\mathbf{X} = (X_1 , X_2, ..., X_p)^T$ , with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Sigma}$ . However, we have to be **careful with perfect correlated variables**, so I will remove the following variables:

```{r}
data_EFA_complete <- data %>%
  dplyr::select(
    -total_sleep_secs,  
    -distance,          
    -calories_total,
    -had_real_activity,
    -activity_status,
    #-stress_avg,
    #-steps,
    #-calories_active
    )
data_EFA <- data_EFA_complete %>%
    dplyr::select(where(is.numeric)) %>%
    na.omit()
```

I will continue using the function fa(). because:

-   factanal (default method is "mle" maximum likelihood estimation. we should only use it if we know that we have normal distribution)

-   fa is non parametric, it is not assuming about normality so we can use it always

**Checking suitability for EFA :**

-   Barlett´s Test of Sphericity:

```{r}
n_total_rows <- nrow(data_EFA)
cor_matrix <- cor(data_EFA)
bartlett_result <- psych::cortest.bartlett(cor_matrix, n = n_total_rows)
print(bartlett_result)
```

According to this test, as the p.value is 0 we can perform EFA.

-   KMO Test:

```{r}
KMO_result <- psych::KMO(cor_matrix)
overall_msa_value <- KMO_result$MSA
cat(paste0("Overall MSA: ", round(overall_msa_value, 3)))
```

```{r}
msa_table <- as.data.frame(KMO_result$MSAi)
colnames(msa_table) <- "MSA"
print(msa_table, digits = 3)
```

According to the KMO Tests, the data is **acceptable** (MSA $\approx$ 0.7) to do EFA.

**How many factors should we choose?**

```{r}
fa_parallel_result <- psych::fa.parallel(data_EFA, 
fm = "pa", #Principal Axis Factoring
fa = "fa")    
```

We apply a rotation so the factors are uncorrelated and maximize the variance of the loadings. This helps interpret the result so variable loads are high on one factor and small on others.

```{r}
num_factors = 5

data_EFA <- data_EFA %>%
  dplyr::select(where(is.numeric)) %>%
  na.omit() %>%
  dplyr::select(where(~ sd(., na.rm = TRUE) > 0)) 

fa_result <- psych::fa(data_EFA, fm = "pa",  nfactors = num_factors, rotate = "varimax")
fa_result
```

-   h2 -\> **communality:** $$h^2_j = \sum_{k=1}^{m} \lambda_{jk}^2$$
-   u2 -\> **uniqueness:** $$u^2_j = 1 - h^2_j$$

Let´s see in more detail the loadings, grouped to interpret the hidden factors in the data:

```{r}
print(fa_result$loadings, cut = 0.4, sort = TRUE)
```

**Factor 1 – Sleep Quality**

-   **The factor increases** when `light_sleep` and `rem_sleep` increase. and the Body battery recharges well.

-   **The factor decreases** when average stress is high.

**Factor 2 - High Intensity Performance**

-   **The factor increases** when activity intensity peaks (`vigorous_secs`, `hr_max`).

-   **The factor increases** along with active calorie burn.

-   **The factor represents** "Sport" rather than general movement.

**Factor 3 – Daily Step Volume**

-   **The factor increases** largely based on `steps` and hitting your `step_goal`.

-   **The factor decreases** when `sport_exposure` increases.

**Factor 4 – Stress**

-   **The factor increases** with the stress, along with the resting heart rate (`rhr`) and minimum heart rate (`hr_min`).

**Factor 5 – Energy reserves**

-   **The factor increases** when the minimum body battery (`bb_min`) stays high.

-   **The factor represents** ending the day with energy left in the tank, rather than draining the battery to zero.

```{r}
colnames(fa_result$loadings) <- c("Sleep\nQuality", 
                "High Intensity\nPerformance", 
                "Stress", 
                "Daily Step\nVolume", # por alguna razon esta al reves
                "Energy Reserves")

psych::fa.diagram(fa_result, cut = 0.4, simple = FALSE,cex = 0.8,l.cex = 0.6, e.size = 0.051,rsize = 0.8,main = "Factor Analysis of Health Metrics (k = 5)") 
```

------------------------------------------------------------------------

# Clustering over PCA data (k=3)

```{r}
pca_data_for_clustering <- rbind(
  train_pca[, 1:3], 
  test_pca[, 1:3]
)
colnames(pca_data_for_clustering) <- c("Recovery vs. Stress","High intensity and caloric burn","Low vs. high intensity exercise")
```

```{r}
numeric_cols <- names(pca_data_for_clustering)[sapply(pca_data_for_clustering, is.numeric)]
for (col in numeric_cols) {
  
  p <- ggplot(pca_data_for_clustering, aes(x = "", y = .data[[col]])) + 
    geom_boxplot() + 
    labs(y = col, title = paste("Boxplot of", col)) + 
    theme_minimal()
  print(p)
}
```

We need to remove the outliers. SPOILER: This takes out a lot of data, so I will not do that, I will cap the max values (I don´t know how legal it is)

```{r}
winsorize <- function(x) {
  lower_limit <- quantile(x, 0.05, na.rm = TRUE)
  upper_limit <- quantile(x, 0.95, na.rm = TRUE)
  pmax(lower_limit, pmin(x, upper_limit))
}
data_capped <- pca_data_for_clustering %>%
  mutate(across(where(is.numeric), winsorize))
```

```{r}
data_scaled <- scale(data_capped)#scale(pca_data_for_clustering)
```

```{r}
# Elbow Method

wcss <- numeric(10)  
for (i in 1:10) {
  kmeans_result <- kmeans(data_scaled, centers = i)
  wcss[i] <- kmeans_result$tot.withinss
}

plot(1:10, wcss, type = "b", pch = 19, col = "blue",
     xlab = "Number of clusters", ylab = "WCSS",
     main = "Elbow Method")
```

### With 4 clusters:

```{r}
num_clusters <- 4
kmeans_result <- kmeans(data_scaled, centers = num_clusters)
print(kmeans_result)
```

```{r}
silhouette_scores <- silhouette(kmeans_result$cluster, dist(data_scaled))
hist(silhouette_scores) # ALGO RARO ESTA OCURRIENDO!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! COnsultar
```

```{r}
avg_silhouette_score <- mean(silhouette_scores[, 3])
print(avg_silhouette_score)
```

```{r}
hopkins_stat <- hopkins(data_scaled)
print(hopkins_stat)
```

```{r}
hopkins.pval(hopkins_stat, nrow(data_scaled))
```

```{r}
# install.packages("factoextra")
library(factoextra)

fviz_cluster(kmeans_result, 
             data = data_scaled, 
             geom = "point",    
             ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "Cluster Visualization (PCA Map)"
)
```

```{r}
library(plotly)
data_scaled = data.frame(data_scaled)
colnames(data_scaled) <- c("Recovery vs. Stress","High intensity and caloric burn","Low vs. high intensity exercise")

plot_ly(data = data_scaled, 
        x = ~`Recovery vs. Stress`,             
        y = ~`High intensity and caloric burn`,   
        z = ~`Low vs. high intensity exercise`,   
        type = "scatter3d",
        mode = "markers",
        color = as.factor(kmeans_result$cluster)) %>% 
  layout(
    title = "3D Cluster Visualization on PCA Data",
    scene = list(
      xaxis = list(title = "Recovery vs. Stress"),
      yaxis = list(title = "High intensity/Caloric burn"),
      zaxis = list(title = "Low vs. High Intensity")
    )
  )
```

```{r}

original_data_combined <- rbind(
  train_data[, -col_remov_number], 
  test_data[, -col_remov_number]
)
original_data_combined$Cluster <- as.factor(kmeans_result$cluster)

cluster_profiling <- original_data_combined %>%
  group_by(Cluster) %>%
  summarise(across(where(is.numeric), mean, .names = "mean_{.col}"))

print(cluster_profiling)
```